The effect of prenatal environment on brain metabolic function and perception in a lizard

### 2025-06-01 ms.qmd
updates

_Oxidative Damage_: Assays of oxidative damage from cryopreserved samples were performed **5 months** after the initial processing and analysis of fresh samples. On the day of oxidative damage assays, we rapidly thawed frozen samples, removed the cryopreservation solution (see above), and resuspended samples in 1 mL warm 1X Tris-EDTA. Afterward, the Tris-EDTA was removed, and samples were resuspended in 200 µL of warm 1X PBS containing 10 µg/mL Hoechst 33342 Nuclear Viability Dye (for cell viability) and 100 µM BODIPY 665/676 Lipid Peroxidation Sensor (to measure lipid peroxidation). Following staining, we permeabilized the cell membranes incubating the samples in 200 µL warm 1X PBS containing 20 µM digitonin. Following permeabilization, we stained the samples with 20 µL of 70 µM 8-OHdG Polyclonal Antibody to measure oxidative damage on DNA. We left the homogenate overnight (~12 hours) at 4 ºC. The following day we counterstained the cells with 20 µL of 100 µg/mL H+G Goat Anti-Rabbit Conjugate Antibody with Alexa-Fluor 488 and analyze the samples in the flow cytometer.

@@

#### Metabolic processes in multiple brain regions are robust to early life conditions
Contrary to our predictions, we found no significant effects of early environmental conditions on mitochondrial density, metabolic capacity, or ROS production in either of the brain regions studied. **EXAMPLES OF OTHER EXPERIMENTS WITH TEMP AND CORT IN THE BRAIN.** Mitochondrial potential is expected to enhance ATP production per oxygen consumed, thereby increasing metabolic efficiency [@rice2002brain; @martinez2016tca]. Given the brain's high energy demands [@mcnay_decreases_2000; @mann_coupling_2021], it is plausible that mitochondria are particularly well-protected, which may help explain their resilience towards environmental inputs.

@@

CORT exposure reduced DNA damage in optic tecta from cold-incubated lizards but had no effect at higher incubation temperatures. This pattern suggests that CORT might play a protective role under specific thermal conditions. Since CORT helps mobilize energy resources to facilitate physiological adjustments to stress [@sapolsky_how_2000], it is possible that under colder incubation conditions — where metabolic rates are expected to be lower — energy mobilized by CORT may be allocated towards antioxidant activity or DNA repairing mechanisms [**REFs**]. In contrast, at higher temperatures, the increased metabolic demands may limit the availability of resources for such protective mechanisms [**REFs**]. **A further possibility is that, under lower metabolic activity induced by cold temperatures more portion of DNA is protected within nucleosomes/histones, decreasing the effects of CORT on oxidative damage** [**REFs**].

In comparison, increased prenatal CORT resulted in higher lipid peroxidation in cold incubated lizards but reduced it in those incubated at warm temperatures. Although the two hormonal treatments did not significantly differ within each incubation temperature, the direction of the CORT effect was the opposite between temperatures. The opposing temperature-dependent effects of CORT on DNA damage and lipid peroxidation indicates that the pathways involved in protection or repair are target-specific. **EXAMPLE OF ENZYMES WORKING ON LIPID PEROXIDATION AND DNA DAMAGE AT DIFFERENT TEMPERATURES**. An alternative mechanism might involve temperature-dependent changes in membrane lipid composition or fluidity, which may affect how lipids respond to oxidative stress [@zeis2019temperature]. For instance, lipid peroxidation levels increased in cold-acclimated _Daphnia magna_ likely because of higher concentration of polyunsaturated fatty acids [@zeis2019temperature]. In that case, temperature-induced structural differences may influence the vulnerability of different molecules to oxidative stress.

#### Prenatal CORT exposure enhances chemical but not visual prey detection   
Detection latency decreased with CORT exposure when animals were presented with chemical, but not visual, stimuli. This effect was not related to alterations in mitochondrial physiology in the olfactory bulbs. Instead, prenatal CORT may influence other mechanisms involved in chemoreception. For instance, in rats (_Rattus norvegicus_), GCs like dexamethasone can enchance the ability to detect chemical stimuli by upregulating gene expression in the olfactory mucosa [@meunier2020olfactory]. Such molecular changes could alter chemosensory sensitivity in lizards, independent of mitochondrial function.

@@

#### Mitochondrial function and oxidative stress do not affect detection abilities
Mitochondria are predicted to impact cognitive performance by providing energy for neural activity or by regulating oxidative stress [@siegel1994basic; @mcnay_decreases_2000; @du_dynamic_2009; @picard_mitochondria_2014; @picard_energetic_2018; @mann_coupling_2021; @alexandrov_neuronal_2022]. However, despite the relevance of metabolic function and oxidative damage on cognitive abilities in other taxa [@zhu_prenatal_2004; @cao2014ampk; @hara_presynaptic_2014] they did not explain variation in prey detection in _L. delicata_. Metabolic function and oxidative damage did not have substantial effects on prey detection in _L. delicata_.

Previous studies have found that **EXAMPLE ABOUT MIT DENSITY OR SIMILAR AND PREY DETECTION IF I CAN FIND ANY BECAUSE DAN DOES NOT REALIZE THAT NOONE DOES THIS**. In contrast, our results suggests that metabolic function is not a limiting factor for prey detection in _L. delicata_, or that _L. delicata_ can sustain this cognitive function despite fluctuations in energy availability.

Similarly, oxidative stress did not influence _L. delicata_ ability to detect visual or chemical stimuli from prey. In this regard, oxidative damage is expected to decrease cognitive abilities through neuron death or senescense [@finkel_oxidants_2000; @zhu_prenatal_2004; @du_dynamic_2009; @gong_chronic_2011; @hoffmann_mitochondrion_2018]. Possibly the levels of oxidative damage observed in our study were not sufficient to impair prey detection either because of compensatory mechanisms - such as antioxidant activity - that counteracted the negative effects of oxidative stress, or because the studied brain regions had a high neural density that provided functional resilience [@rice2002brain; @amiel_smart_2011]. Alternatively, the cognitive effects of oxidative stress may not yet be detectable at this stage. Oxidative damage accumulates over time, and its effects on cognition can pronounce with age [@terman2006oxidative; @hara_presynaptic_2014]. In fact, we saw a significant effect of age on DNA damage in olfactory bulbs, and on lipid peroxidation in both regions. Cross-sectional studies across life stages ay help reveal how oxidative damage shapes cognitive function over time.

@@

**Staining cryopreserved samples**: Assays of oxidative damage from cryopreserved samples were performed within 6 months of the initial processing and analysis of fresh samples. On the day of oxidative damage assays, we rapidly thawed frozen samples by briefly (1-2 minutes) submerging them in warm water. We washed each thawed sample twice, the first time resuspending the pelleted cells in 1 mL warm 1X Tris-EDTA, and the second time in 200 µL warm 1X PBS. Pooled homogenate was divided into 100 µL aliquots to serve as single-color controls. Thawed, resuspended samples were first stained with 20 µL (10 µL for controls) of #µM Hoechst 33342 Nuclear Viability Dye and 20 µL of 100 µM BODIPY 665/676 Lipid Peroxidation Sensor and incubated at 32 °C for 20 minutes. Following staining, we washed cells to prevent further binding of unbound fluorescent probes, then resuspended the pellet in 200 µL warm 1X PBS containing 20 µM digitonin. We incubated the samples at 32 °C for 20 minutes to permeabilize the cell membrane, after which we washed the homogenate and resuspended the pelleted cells in 200 µL 1X PBS. We added 20 µL of 70 µM 8-OHdG Polyclonal Antibody to each sample, after which we left the homogenate overnight (~12 hours) for the antibody to bind to 8-OHdG, a marker of oxidative damage on DNA. The following day we counterstained the cells with 20 µL of #µM (100 µg/mL) H+G Goat Anti-Rabbit Conjugate Antibody with Alexa-Fluor 488 at 32 °C for 20 minutes. After the cells had been tagged with 8-OHdG antibodies and counterstained, we washed the cells once more and resuspended the pellet in 400 µL of 1X PBS. Unstained and single-color controls were treated identically to samples, but stained with only up to one of BODIPY 665/676, Hoechst 33342, 8-OHdG antibody, or Alexa-Fluor 488 conjugate. Additionally, one control was stained with both 8-OHdG antibody and the Alexa-Fluor 488 conjugate. We then loaded a 96-well plate with 100 µL of each single-color control and 100 µL in duplicate of each sample. We performed all flow cytometry assays on samples within 48-hours of thawing the samples.

>>>>

_Oxidative Damage_: Assays of oxidative damage from cryopreserved samples were performed 61 weeks after the initial processing and analysis of fresh samples. On the day of oxidative damage assays, we rapidly thawed frozen samples, removed the cryopreservation solution (see above), and resuspended samples in 1 mL warm 1X Tris-EDTA. Afterward, the Tris-EDTA was removed, and samples were resuspended in 200 µL of warm 1X PBS containing 10 µg/mL Hoechst 33342 Nuclear Viability Dye (for cell viability) and 100 µM BODIPY 665/676 Lipid Peroxidation Sensor (to measure lipid peroxidation). Following staining, we permeabilized the cell membranes incubating the samples in 200 µL warm 1X PBS containing 20 µM digitonin. Following permeabilization, we stained the samples with 20 µL of 70 µM 8-OHdG Polyclonal Antibody to measure oxidative damage on DNA. We left the homogenate overnight (~12 hours) at 4 ºC. The following day we counterstained the cells with 20 µL of 100 µg/mL H+G Goat Anti-Rabbit Conjugate Antibody with Alexa-Fluor 488 and analyze the samples in the flow cytometer.

@@

#### Metabolic processes in multiple brain regions are robust to early life conditions
Contrary to our predictions, we found no significant effects of early environmental conditions on mitochondrial density, metabolic capacity, or ROS production in either of the brain regions studied. In the leopard gecko (_Eublepharis macularius_), incubation temperature is an important determinant of metabolic capacity in some, but not all, brain nuclei [@coomber_independent_1997], while prenatal stress decreases mitochondrial respiration efficiency in the cortex, hippocampus, and hypotalamus of rats (_Rattus norvegicus_) [@gong_chronic_2011]. In contrast, metabolic processes in the optic tecta and olfactory bulbs of _L. delicata_ appear to be robust to early-life conditions, potentially to ensure that these brain regions can perform their essential functions. However, since the effects of early environment can be region-dependent [@coomber_independent_1997], it is possible that other brain regions not examined here may be more sensitive to prenatal CORT and temperature.

@@

CORT exposure reduced DNA damage in optic tecta from cold-incubated lizards but had no effect at higher incubation temperatures. This pattern suggests that CORT might play a protective role under specific thermal conditions. Since CORT helps mobilize energy resources to facilitate physiological adjustments to stress [@sapolsky_how_2000], it is possible that under colder incubation conditions — where metabolic rates are expected to be lower — energy mobilised by CORT may be allocated towards antioxidant activity or DNA repairing mechanisms [@kawamura1991glucocorticoid; but see @mcintosh1998glucocorticoids]. In contrast, at higher temperatures, the increased metabolic demands may limit the availability of resources for such protective mechanisms [see @kim2019carry].

In comparison, increased prenatal CORT resulted in higher lipid peroxidation in cold incubated lizards but reduced it in those incubated at warm temperatures. Although the two hormonal treatments did not significantly differ within each incubation temperature, the direction of the CORT effect was the opposite between temperatures. These contrasting CORT–temperature effects on DNA damage and lipid peroxidation suggest that the pathways protecting or repairing each molecule may vary in their thermal sensitivity. For instance, antioxidant enzymes differ in their response to hibernation in the soft-shelled turtle (_Pelodiscus sinensis_), indicating that thermal sensitivity varies among antioxidant system components [@tang2021antioxidant]. An alternative mechanism might involve temperature-dependent changes in membrane lipid composition or fluidity, which may affect how lipids respond to oxidative stress [@zeis2019temperature]. For example, lipid peroxidation levels increased in cold-acclimated _Daphnia magna_ likely because of higher concentration of polyunsaturated fatty acids [@zeis2019temperature]. Therefore, temperature-induced structural differences may influence the vulnerability of different molecules to oxidative stress.

#### Prenatal CORT exposure enhances chemical but not visual prey detection   
Detection latency decreased with CORT exposure when animals were presented with chemical, but not visual, stimuli. This effect was not related to alterations in mitochondrial physiology in the olfactory bulbs. Instead, prenatal CORT may influence other mechanisms involved in chemoreception. For instance, in rats (_R. norvegicus_), GCs like dexamethasone can enchance the ability to detect chemical stimuli by upregulating gene expression in the olfactory mucosa [@meunier2020olfactory]. Such molecular changes could alter chemosensory sensitivity in lizards, independent of mitochondrial function.

@@

#### Mitochondrial function and oxidative stress do not affect detection abilities
Mitochondria are predicted to impact cognitive performance by providing energy for neural activity or by regulating oxidative stress [@siegel1994basic; @mcnay_decreases_2000; @du_dynamic_2009; @picard_mitochondria_2014; @picard_energetic_2018; @mann_coupling_2021; @alexandrov_neuronal_2022]. For example, reduced cytochrome c oxidase (COX) activity in dopaminergic neurons of the olfactory bulb of mice (_Mus musculus_) is associated with impaired olfactory capacities [@pass2020impact], and links between brain metabolic function and visual acuity have also been reported in other species [reviewed in @wong2010energy]. However, metabolic function did not explain variation in prey detection in _L. delicata_. In contrast, our results suggests that metabolic function is not a limiting factor for prey detection in _L. delicata_, or that _L. delicata_ can sustain this cognitive function despite fluctuations in energy availability.

Similarly, oxidative stress did not influence _L. delicata_ ability to detect visual or chemical stimuli from prey. In this regard, oxidative damage is expected to decrease cognitive abilities through neuron death or senescense [@finkel_oxidants_2000; @zhu_prenatal_2004; @du_dynamic_2009; @gong_chronic_2011; @hoffmann_mitochondrion_2018]. Possibly, the levels of oxidative damage observed in our study were not sufficient to impair prey detection either because of compensatory mechanisms - such as antioxidant activity - that counteracted the negative effects of oxidative stress, or because the studied brain regions had a high neural density that provided functional resilience [@rice2002brain; @amiel_smart_2011]. Alternatively, the cognitive effects of oxidative stress may not yet be detectable at this stage. Oxidative damage accumulates over time, and its effects on cognition can pronounce with age [@terman2006oxidative; @hara_presynaptic_2014]. As such, we could find a strong relationship if older individuals were employed, but not in lizards this age. In fact, we saw a significant effect of age on DNA damage in olfactory bulbs, and on lipid peroxidation in both regions. Cross-sectional studies across life stages may help reveal how oxidative damage shapes cognitive function over time.

@@

**Staining cryopreserved samples**: Assays of oxidative damage from cryopreserved samples were performed 61 weeks after the initial processing and analysis of fresh samples. On the day of oxidative damage assays, we rapidly thawed frozen samples by briefly (1-2 minutes) submerging them in warm water. We washed each thawed sample twice, the first time resuspending the pelleted cells in 1 mL warm 1X Tris-EDTA, and the second time in 200 µL warm 1X PBS. Pooled homogenate was divided into 100 µL aliquots to serve as single-color controls. Thawed, resuspended samples were first stained with 20 µL (10 µL for controls) of #µM Hoechst 33342 Nuclear Viability Dye and 20 µL of 100 µM BODIPY 665/676 Lipid Peroxidation Sensor and incubated at 32 °C for 20 minutes. Following staining, we washed cells to prevent further binding of unbound fluorescent probes, then resuspended the pellet in 200 µL warm 1X PBS containing 20 µM digitonin. We incubated the samples at 32 °C for 20 minutes to permeabilize the cell membrane, after which we washed the homogenate and resuspended the pelleted cells in 200 µL 1X PBS. We added 20 µL of 70 µM 8-OHdG Polyclonal Antibody to each sample, after which we left the homogenate overnight (~12 hours) for the antibody to bind to 8-OHdG, a marker of oxidative damage on DNA. The following day we counterstained the cells with 20 µL of #µM (100 µg/mL) H+G Goat Anti-Rabbit Conjugate Antibody with Alexa-Fluor 488 at 32 °C for 20 minutes. After the cells had been tagged with 8-OHdG antibodies and counterstained, we washed the cells once more and resuspended the pellet in 400 µL of 1X PBS. Unstained and single-color controls were treated identically to samples, but stained with only up to one of BODIPY 665/676, Hoechst 33342, 8-OHdG antibody, or Alexa-Fluor 488 conjugate. Additionally, one control was stained with both 8-OHdG antibody and the Alexa-Fluor 488 conjugate. We then loaded a 96-well plate with 100 µL of each single-color control and 100 µL in duplicate of each sample. We performed all flow cytometry assays on samples within 48-hours of thawing the samples.
